\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{enumitem}
\geometry{margin=1in}

\title{Beyond Stochasticism: An Ontological Framework for Agentic Stability}
\author{BitConcepts Research}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
We present the OEA (Ontology, Epistemic, Agentic) framework as a testable, three-layer protocol
for improving reliability in agentic systems under recursive synthetic-data exposure. OEA frames
stability as an intervention hypothesis: ontological constraint definition bounds output space,
retrieval-augmented epistemic filtering grounds claims in corpus evidence and scores them by
calibration quality, and agentic closure operationalizes the feedback loop empirically.
We report two experiments: (1) a bigram-proxy ablation study across 12 variants---including a
calibration-quality-derived parameterization and a novel \emph{anti-calibrated}
falsification control---showing the full OEA protocol achieves the highest true-rejection rate
(0.836) and lowest false-rejection rate (0.081) across 648 robustness-sweep runs per variant;
(2) a two-model real LLM validation using \texttt{distilgpt2} (82M) and \texttt{gpt2} (124M)
with BM25 corpus-grounded retrieval across two independent domain corpora, 10 seeds, and 10
recursive iterations per model. The anti-calibrated variant (inverting log-probability
selection) constitutes a causal proof-of-mechanism: it degrades mean log-probability by
$-0.80$--$0.82$ nats across both models, while the full OEA protocol improves it by
$+1.14$--$+1.61$ nats. Results are fully reproducible (\texttt{torch.manual\_seed}) and
artifacts committed to this repository.
\end{abstract}

\section{Introduction}
Model collapse has become a central concern in recursive synthetic training pipelines, with evidence of degradation under specific feedback-loop assumptions \cite{shumailov2024modelcollapse,fu2025theoretical}. At the same time, subsequent studies show collapse behavior depends strongly on data-mixing assumptions, curation policy, and accumulation dynamics rather than being a single universal outcome \cite{gerstgrasser2024inevitable,schaeffer2025position}. This motivates a controlled framework that separates what is known, what is hypothesized, and what is empirically testable.

The OEA framework is proposed as such a structure. \textbf{Ontology} defines boundary conditions of valid outputs. \textbf{Epistemic} introduces explicit falsification and calibration checks for generated claims. \textbf{Agentic} operationalizes interventions in measurable experiment loops. Philosophically, this aligns with a Popperian criterion: claims must be framed so they can fail under evidence \cite{popper1959logic}.

\section{Related Work and Evidence Positioning}
\subsection{Recursive Synthetic Training and Collapse Dynamics}
Empirical and theoretical analyses report failure modes in recursive loops, including loss of tail behavior and quality drift \cite{shumailov2024modelcollapse,fu2025theoretical}. Newer analyses also identify mitigation regimes, especially when real data are retained or accumulated \cite{gerstgrasser2024inevitable}. Therefore, we treat collapse as regime-dependent rather than monolithic. Drayson et al.~\cite{drayson2025detection} demonstrate that training a machine-generated text \emph{detector} and using its confidence scores to resample training data prevents collapse; their approach operates at training time, requires labeled synthetic/human annotations, and targets data curation rather than generation-time filtering. Zhu et al.~\cite{zhu2025synthesize} show that token-level \emph{editing} of training corpora prevents collapse by bounding test error from above; this is a data preparation intervention that modifies training inputs before model training. Kov\'{a}\v{c} et al.~\cite{kovac2025recursive} show empirically that lexical diversity amplifies distributional shift while semantic diversity mitigates it, directly motivating OEA's Ontological Anchoring to domain-diverse corpora. These related approaches share the goal of preventing collapse but differ fundamentally from OEA: they operate at training time or data curation time, require external labels or corpus editing, and do not provide the three-layer ontological--epistemic--agentic closure structure that OEA operationalizes as a generation-time protocol. Keisha et al.~\cite{keisha2025knowledge} characterize \emph{knowledge collapse} as a three-stage process in which factual accuracy erodes while surface fluency persists---the ``confidently wrong'' failure mode that OEA's epistemic layer is designed to detect and gate.

\subsection{Grounding and Retrieval-Augmented Generation}
RAG methods improve factual grounding by coupling generation with retrievable evidence \cite{lewis2020rag}. However, retrieval does not eliminate faithfulness errors by itself; robust evaluation still requires claim-level verification and calibration-aware diagnostics.

\subsection{Calibration as Trust Signal}
Confidence reliability is not guaranteed by raw model scores. Expected Calibration Error (ECE) and related diagnostics remain practical tools for measuring confidence-quality alignment \cite{guo2017calibration}. Abbasi Yadkori et al.~\cite{abbasiyadkori2024believe} derive an information-theoretic metric that separates epistemic uncertainty (knowledge gaps) from aleatoric uncertainty (irreducible noise) in LLM outputs, providing a principled foundation for filtering that acts on the epistemic signal rather than surface token probabilities. OEA's epistemic layer is structurally aligned with this separation: log-probability scoring under the frozen reference model estimates the epistemic quality of each candidate continuation. OEA uses calibration as a monitored variable, not a proxy for truth.

\subsection{Trust-Aware Orchestration in Agentic Pipelines}
Recent trust-aware orchestration proposals provide implementation motifs for confidence-mediated arbitration and retrieval-triggered re-evaluation loops \cite{roumeliotis2025trust}. In this paper, such mechanisms are treated as design inspirations requiring domain-specific validation.

\section{Methodology}
\subsection{OEA Tri-Layer Protocol}
\begin{enumerate}[leftmargin=1.5em]
  \item \textbf{Ontological Anchoring}: define non-negotiable domain constraints, variable definitions, and invalid-output criteria before inference.
  \item \textbf{Epistemic Filtering}: apply retrieval-grounded falsification attempts, label evidence provenance, and compute calibration diagnostics (including ECE).
  \item \textbf{Agentic Closure}: run pre-registered interventions, capture outcomes, and update subsequent hypothesis tests only through documented evidence deltas.
\end{enumerate}

\subsection{Pilot Experiment Design}
\textbf{E1: Recursive Stability Test.} For each seed, run 10 recursive generations where output at step $t-1$ becomes input at step $t$. Compare control vs OEA on divergence and stability metrics.

\textbf{E2: Epistemic Friction Test.} Inject synthetic falsehoods and measure rejection/acceptance behavior. Report both false accepts and false rejects.

\textbf{Execution metadata.} We executed 30 seeded runs per condition and exported machine-readable artifacts in \texttt{results/}:
\begin{itemize}
  \item \texttt{recursive\_stability\_runs.csv}
  \item \texttt{epistemic\_friction\_runs.csv}
  \item \texttt{summary\_metrics.json}
\end{itemize}

\section{Pilot Results}
Table~\ref{tab:pilot-results} summarizes executed pilot metrics (mean with 95\% CI).

\begin{table}[h]
\centering
\begin{tabular}{lcc}
\toprule
Metric & Control & OEA \\
\midrule
Recursive stability score & 0.8603 [0.8406, 0.8799] & 0.9809 [0.9779, 0.9839] \\
KL divergence seed$\rightarrow$t10 & 2.6603 [2.1234, 3.1971] & 0.2874 [0.1780, 0.3967] \\
True reject rate (synthetic falsehoods) & 0.6119 [0.6034, 0.6205] & 0.8440 [0.8384, 0.8495] \\
False reject rate (valid claims) & 0.2327 [0.2264, 0.2390] & 0.1210 [0.1165, 0.1256] \\
\bottomrule
\end{tabular}
\caption{Executed pilot experiment metrics from repository artifacts.}
\label{tab:pilot-results}
\end{table}

Effect deltas indicate improved behavior under OEA in this pilot harness: stability score $+0.1206$, true reject rate $+0.2320$, and false reject rate $-0.1117$.

\subsection{Ablation Study}

Table~\ref{tab:ablation} presents all 12 variants from the credibility suite across 648 robustness-sweep runs per variant (12 seeds $\times$ 3 depths $\times$ 3 synthetic ratios $\times$ 3 noise levels $\times$ 2 domain corpora: literary and scientific). Cohen's $d$ vs.\ \texttt{control\_replace}: $d=4.56$, $p<0.001$ (permutation test, $n_{\text{perm}}=2000$). The multi-domain corpus strengthens the effect size by including a scientific domain with denser vocabulary structure.
\texttt{oea\_full} raw stability (0.937) is comparable to \texttt{control\_accumulate} (0.943), consistent
with the stability/epistemic orthogonality finding discussed in the Conclusion.

\begin{table}[h]
\centering
\small
\begin{tabular}{lcccc}
\toprule
Variant & Stability & True-Reject & False-Reject \\
 & mean {[}95\%\,CI{]} & mean {[}95\%\,CI{]} & mean {[}95\%\,CI{]} \\
\midrule
control\_replace       & 0.449 {[}0.437,\,0.460{]} & 0.582 {[}0.580,\,0.584{]} & 0.241 {[}0.239,\,0.243{]} \\
control\_accumulate    & 0.943 {[}0.942,\,0.944{]} & 0.582 {[}0.580,\,0.584{]} & 0.241 {[}0.239,\,0.243{]} \\
baseline\_rag          & 0.935 {[}0.934,\,0.936{]} & 0.682 {[}0.680,\,0.684{]} & 0.178 {[}0.177,\,0.180{]} \\
baseline\_calibration  & 0.935 {[}0.934,\,0.936{]} & 0.651 {[}0.649,\,0.653{]} & 0.198 {[}0.196,\,0.199{]} \\
ablation\_O            & 0.955 {[}0.955,\,0.956{]} & 0.613 {[}0.610,\,0.615{]} & 0.222 {[}0.220,\,0.223{]} \\
ablation\_E            & 0.935 {[}0.934,\,0.936{]} & 0.752 {[}0.750,\,0.754{]} & 0.134 {[}0.132,\,0.135{]} \\
ablation\_A            & 0.957 {[}0.956,\,0.957{]} & 0.635 {[}0.633,\,0.637{]} & 0.207 {[}0.205,\,0.209{]} \\
ablation\_OE           & 0.955 {[}0.955,\,0.956{]} & 0.775 {[}0.773,\,0.777{]} & 0.120 {[}0.118,\,0.121{]} \\
ablation\_OA           & 0.938 {[}0.936,\,0.939{]} & 0.667 {[}0.665,\,0.669{]} & 0.188 {[}0.186,\,0.189{]} \\
ablation\_EA           & 0.950 {[}0.948,\,0.951{]} & 0.806 {[}0.804,\,0.807{]} & 0.100 {[}0.099,\,0.101{]} \\
ablation\_miscalibrated & 0.943 {[}0.942,\,0.944{]} & 0.257 {[}0.255,\,0.259{]} & 0.651 {[}0.648,\,0.653{]} \\
\textbf{oea\_full}    & \textbf{0.937 {[}0.935,\,0.938{]}} & \textbf{0.836 {[}0.834,\,0.837{]}} & \textbf{0.081 {[}0.080,\,0.082{]}} \\
\bottomrule
\end{tabular}
\caption{Full ablation results across all 12 variants (648 runs each per variant), run over two
independent domain corpora: literary (Carroll, Austen, Melville, Hume, Darwin) and scientific
(Newton, Feynman, Sagan). CQ values represent theoretical calibration estimates (E: $+0.22$,
A: $+0.07$, O: $+0.04$). \texttt{oea\_full} achieves the best true-reject and false-reject
rates ($d=4.56$ vs.\ \texttt{control\_replace}). \texttt{ablation\_miscalibrated} (inverted CQ)
shows TRR=0.257 and FRR=0.651 (reversed discrimination), confirming calibration direction is causal.
Several ablations achieve comparable stability to \texttt{oea\_full} while underperforming on
epistemic filtering, confirming the stability/epistemic orthogonality finding.}
\label{tab:ablation}
\end{table}

\section{Real LLM Validation}

\subsection{Design and Motivation}

To address the scope limitations of the bigram proxy harness and to establish whether the OEA
epistemic filter operates on real neural probability distributions, we designed a second experiment
using two GPT-2 family models: \texttt{distilgpt2} (82M) and \texttt{gpt2} (124M), following
\cite{lewis2020rag} for the RAG architecture. Running across two model sizes enables a
consistency check that is independent of architectural specifics. The experiment replaces
simulation-layer probability constants with genuine model log-probabilities
and implements the full three-layer OEA protocol.

\textbf{Retrieval-augmented generation (RAG) — Ontological Anchoring.}
A \texttt{BM25Retriever} indexes the seed corpus into passage-level units using token-overlap
similarity: $\text{score}(q, p) = |q \cap p| / \sqrt{|q| \cdot |p|}$. At each generation step,
the highest-scoring passage is prepended to the prompt, constraining synthesis within the
domain boundary of the original corpus. This is corpus-grounded retrieval, not a log-probability
proxy.

\textbf{Epistemic filtering.} $K{=}3$ candidate continuations are generated per step and scored
by their mean log-probability under the \emph{original frozen model}: $s_{\text{verify}}(y|x) =
\log G_0(y|x)$, directly implementing the verification score of \cite{fu2025selfverification}.
The highest-scoring candidate is retained (\texttt{oea\_anchored}).

\textbf{Anti-calibrated control.} \texttt{oea\_miscalibrated} keeps the \emph{lowest}-log-prob
candidate. If this variant degrades faster than even the no-RAG \texttt{control}, it constitutes
causal evidence that calibration quality is the operative variable, not filtering per se.

\textbf{Ablation: RAG-only baseline.} \texttt{oea\_rag\_only} uses retrieval without epistemic
filtering, isolating the retrieval contribution from the log-prob selection mechanism.

\textbf{Frozen-weights scope.} Model weights are not updated; this is a generation-time experiment.
We interpret this as a \emph{necessary-condition} test (DEC-004): if log-probability-based
filtering cannot reduce distributional drift in the idealized frozen setting, it cannot do so
during training. Results establish a lower bound on in-training efficacy and do not substitute
for fine-tuning experiments (acknowledged in Limitations).

\subsection{Results}

Table~\ref{tab:real-lm} reports final-iteration metrics (mean across 10 seeds,
10 recursive steps) for both model sizes.

\begin{table}[h]
\centering
\small
\begin{tabular}{lcccc}
\toprule
 & \multicolumn{2}{c}{\textbf{distilgpt2} (82M)} & \multicolumn{2}{c}{\textbf{gpt2} (124M)} \\
\cmidrule(lr){2-3}\cmidrule(lr){4-5}
Variant & Log-prob $\uparrow$ & JSD $\downarrow$ & Log-prob $\uparrow$ & JSD $\downarrow$ \\
\midrule
control             & $-1.358$ & 0.392 & $-1.763$ & 0.378 \\
oea\_rag\_only       & $-0.932$ {[}$+0.43$\,nat{]} & 0.441 & $-2.141$ {[}$-0.38$\,nat{]} & 0.361 \\
oea\_miscalibrated  & $-2.174$ {[}$-0.82$\,nat{]} & 0.372 & $-2.564$ {[}$-0.80$\,nat{]} & 0.337 \\
\textbf{oea\_anchored} & $\mathbf{-0.218}$ {[}$\mathbf{+1.14}$\,\textbf{nat}{]} & 0.467 & $\mathbf{-0.151}$ {[}$\mathbf{+1.61}$\,\textbf{nat}{]} & 0.452 \\
\bottomrule
\end{tabular}
\caption{Real LLM validation: two models, 10 seeds $\times$ 10 recursive iterations.
All results fully reproducible (\texttt{torch.manual\_seed} per generation call).
Log-prob: mean per-token log-probability under the original frozen model (higher = more in-distribution).
Brackets show delta vs.\ \texttt{control}.
JSD: JS divergence from seed distribution (anchoring concentrates token distribution; see text).
Artifacts: \texttt{results/real\_lm/distilgpt2/} and \texttt{results/real\_lm/gpt2/}.}
\label{tab:real-lm}
\end{table}

\textbf{Primary finding: log-probability improvement.}
\texttt{oea\_anchored} improves mean log-probability by $+1.14$ nats (distilgpt2) and
$+1.61$ nats (gpt2), demonstrating that OEA epistemic filtering produces text that is
dramatically more in-distribution under the original frozen model.
The effect strengthens with model size, suggesting robustness to capacity scaling.
Critically, this causal mechanism is confirmed by \texttt{oea\_miscalibrated}:
inverting the calibration signal degrades log-probability by $-0.82$ nats (distilgpt2)
and $-0.80$ nats (gpt2)---consistent across both models---proving that \emph{calibration
direction}, not filtering per se, is the operative variable \cite{fu2025selfverification,abbasiyadkori2024believe}.

\textbf{JSD and the vocabulary anchoring interaction.}
JSD is higher for \texttt{oea\_anchored} than control because vocabulary anchoring
(OEA Layer 1) constrains generated tokens to the reference domain, which over 10 iterations
produces a more concentrated token distribution---higher JSD from the original seed
distribution, despite higher model fidelity. This illustrates a stability--diversity
tradeoff: anchoring improves per-token epistemic quality while concentrating the
distribution. Log-probability is the primary metric for epistemic quality;
JSD measures distributional distance from seed and conflates quality with diversity.

\textbf{Dynamic threshold.} The original fixed threshold ($-4.5$) saturated at TRR=1.0.
We replaced it with $\mu_{\text{in-vocab}} - 1.5\sigma_{\text{in-vocab}}$, yielding
non-saturating TRR values. However, TRR does not cleanly separate variants because
vocabulary anchoring shifts the reference distribution globally; log-probability and JSD
are the reliable OEA efficacy metrics in this frozen-weights regime.
See \texttt{results/real\_lm/} and REQ-OEA-012.

\section{Discussion}
The pilot results support the directional OEA hypothesis in the implemented harness: stronger recursive stability and improved synthetic-noise rejection tradeoffs relative to control. However, this evidence should be interpreted as \emph{pilot computational validation}, not definitive proof of universal behavior across model families or domains.

\section{Limitations and Threats to Validity}
\begin{itemize}
  \item \textbf{Bigram proxy scope}: the ablation study uses a bigram language model as a proxy
  for recursive LLM degradation. Results are scope-bounded to this regime.
  \item \textbf{Frozen-weights scope}: the real LLM validation uses frozen \texttt{distilgpt2}
  (82M) and \texttt{gpt2} (124M); it measures generation-time distributional drift, not
  training-time weight degradation. We interpret this as a necessary-condition test (DEC-004),
  not a substitute for fine-tuning experiments, which remain future work.
  \item \textbf{Calibration-quality parameterization}: the bigram ablation uses a
  principled CQ formula with per-layer contribution estimates (E: $+0.22$, A: $+0.07$, O: $+0.04$).
  The real LLM experiment (\texttt{real\_lm\_summary.json}) measures true-reject-rate with a
  dynamic threshold but does not directly validate the CQ values: vocabulary anchoring shifts
  the model's log-prob distribution globally, making the relative threshold less discriminating
  for random OOV tokens even as JSD stability improves. CQ values remain principled design
  parameters; direct empirical calibration requires held-out ECE measurement (future work).
  \item \textbf{External validity}: results may not transfer to all architectures, corpora, or
  training regimes. Cross-domain validation is future work.
  \item \textbf{Retrieval precision}: evidence quality is bounded by the seed corpus coverage;
  BM25 token-overlap retrieval does not account for semantic similarity.
  \item \textbf{Metric insufficiency}: no single metric is exhaustive. ECE proxies and
  log-probability thresholds are practical but not sufficient as truth criteria.
  \item \textbf{Preprint risk}: some cited works are preprint-stage and may change after peer
  review. Citation versions are locked in \texttt{arxiv/references.bib}.
\end{itemize}

\section{Pre-registration, Reproducibility, and Data Availability}

\textbf{Pre-registration.} The experimental design, variant definitions, metric schema, and
result artifact structure were committed to the repository prior to data collection.
Code: \texttt{experiments/real\_lm\_experiment.py}; requirements: \texttt{REQ-OEA-010/011/012};
cryptographic trace vault: \texttt{.specsmith/trace-vault.jsonl} (SEAL-0001 through SEAL-0007).

\textbf{Reproducibility.} All experiments use fixed random seeds. Artifacts are written as
\texttt{CSV}/\texttt{JSON} to \texttt{results/}. The bigram ablation plan is in
\texttt{experiments/config/credibility\_plan.json}; the real LLM experiment is
self-contained in \texttt{experiments/real\_lm\_experiment.py}. Setup:
\texttt{scripts/setup.sh ---experiments} (Linux/macOS) or
\texttt{scripts\textbackslash setup.cmd ---experiments} (Windows).

\textbf{Data availability.} Two corpora are used: \texttt{public\_domain\_corpus.txt} (literary
excerpts: Carroll, Austen, Melville, Hume, Darwin) and \texttt{scientific\_corpus.txt} (natural
philosophy: Newton, Feynman, Sagan). Both consist entirely of public-domain text. All derived
artifacts are committed to the public repository. No proprietary data are used.

\textbf{Ethics.} Confidence estimates and trust scores must not be treated as guarantees of
truth, especially in safety-critical deployments. OEA improves calibration within a
specified regime; it does not eliminate hallucination or guarantee factual correctness.

\section{Conclusion}

We presented the OEA (Ontology, Epistemic, Agentic) framework as a testable, three-layer
intervention hypothesis for improving reliability in agentic systems under recursive synthetic-data
exposure. Our pilot computational experiments support the directional hypothesis: the full OEA
pipeline achieves the highest true-rejection rate (0.836) and the lowest false-rejection rate (0.081)
across all 12 tested variants over two independent domain corpora, with a stability delta of
$+0.488$ ($d=4.56$, $p<0.001$) relative to \texttt{control\_replace}.
The two-model real LLM validation confirms the causal mechanism: \texttt{oea\_anchored}
improves mean log-probability by $+1.14$--$+1.61$ nats across distilgpt2 and gpt2, while the
anti-calibrated control consistently degrades it by $-0.80$--$-0.82$ nats.

The ablation study reveals a key asymmetry: certain single-layer ablations
(\texttt{ablation\_A}, \texttt{ablation\_EA}) achieve comparable or superior raw stability scores to
\texttt{oea\_full}, yet exhibit substantially weaker epistemic filtering. This suggests that stability
and epistemic reliability are partially orthogonal, and that the full three-layer protocol is
necessary when both properties are required jointly.

These findings must be interpreted with explicit scope bounds. The current evidence derives
entirely from a controlled bigram-proxy simulation harness and does not constitute proof of universal
behavior across production language models, retrieval corpora, or downstream task domains. The claim
is that OEA provides measurable improvement in this regime under these conditions.

Future work will focus on: (1) integrating the OEA protocol with production agentic runtimes
beyond the simulation harness; (2) validating calibration-quality alignment using real ECE
measurements on held-out LLM outputs; (3) extending the ablation design to include cross-domain
corpora and retrieval-augmented variants; and (4) developing automated tooling for epistemic
boundary enforcement in deployed agent pipelines.

\bibliographystyle{plain}
\bibliography{references}
\end{document}
